Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-19 doc/network.tex
Minor stuff at meeting
JD fiddles abstract

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, provides a link between two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Often, generation intervals are measured through contact tracing by identifying who infected whom;
we study how these observed intervals vary both spatially and temporally.
Temporal variation in generation intervals is caused by the censoring in the observation process through contact tracing, whereas spatial variation in generation intervals reflects the patterns of spread of disease.
Therefore, correcting for temporal variation, but not for spatial variation, provides an ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.

@@

%% SWP: see or cite
%% https://www.ncbi.nlm.nih.gov/pubmed/29087987

@@

%% \dc{you make several assertions that are difficult to understand that early in the text. It really doesn't flow well, it was hard to follow, evn fo me who has worked on this kind of things. I suggest you define first c_t.}
%% \jd{I think that verbal explanations first are better for a larger fraction of people. Maybe we can just make the verbal explanations better. I agree that we should keep the horrible math if DC likes it.}

@@

The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
\end{equation}
As shown in \cite{champredon2015intrinsic},
%% \dc{I added the reference for the reader who wants to see how exactly this is derived}
the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
\begin{equation}\label{eq:backward}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by normalizing
\begin{equation}\label{eq:obsg}
\begin{aligned}
c_t(\tau) 
&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
\end{aligned}
\end{equation}
Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
\end{equation}

@@

Estimating generation-interval distributions is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, local spatial (or network) effects and population-level effects.
We have shown that correcting for temporal effects allows us to estimate a spatially informed distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease. 
%% Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.

\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987). You also did it with Trapman, etc. throughout the main text, but it would be nice to summarize }

@@

To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\jd{Mention that our main-text models use exponential later instead.}
\swp{Done.}
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
While we use the SEIR model for simulations in our main text, we show results based on more realistic distributions in Appendix.

@@

\jd{Add flowchart?}
\swp{Not sure what kind of flowchart you're thinking about.. let me think more...}

>>>>

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Often, generation intervals are measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales. 
We show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.

@@

@@

@@

The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
\end{equation}
As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
\begin{equation}\label{eq:backward}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by normalizing
\begin{equation}\label{eq:obsg}
\begin{aligned}
c_t(\tau) 
&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
\end{aligned}
\end{equation}
Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
\end{equation}

@@

\jd{\P\ about contact tracing.}
\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987). You also did it with Trapman, etc. throughout the main text, but it would be nice to summarize }

Estimating generation-interval distributions is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, local spatial (or network) effects and population-level effects.
We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease. 
%% Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.

@@

To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods. This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}: 
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
While we use the SEIR model for simulations in our main text, we show results based on more realistic distributions in Appendix.

@@
